---
title: "Blog 5: Geographic Analysis of Experiment and Census Data"
description:  |
  Leverage Census data to check the geographical distribution of the experimental design..
author: "TEAMATE"
date: "2025-04-07"
draft: FALSE
---

This code continues the same procedure as the previous blog, focusing on the analysis of a combined state-by-year dataset. 

Note: Experiment data is from 2019-2021, while census data is from 2019-2024. The following code aggregate application level data to the **state** level and merge it with the census data to calculate the proportion of applications by state.

This exercise will help us understand the background of the data on a state-by-state basis and explore the relationship between callback rates and other variables at the state level.

```{r}
rm(list = ls())
# hide warnings
options(warn = -1)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(stargazer))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(fixest))
suppressPackageStartupMessages(library(usmap))


data <- readRDS("dataset/cleaned_data.rds")
census_data <- read_excel('dataset/NST-EST2024-POP.xlsx')
census_data_2 <- read_excel('dataset/nst-est2019-01.xlsx', skip = 3)

# for census_data_2 only keep year column 13 and row 6 onwards
census_data_2 <- census_data_2 %>%
  select(1, 13) %>%
  slice(6:56) 

# rename first column to states 
colnames(census_data_2)[1] <- "states"

# delete . in front of state names
census_data_2 <- census_data_2 %>% 
  filter(grepl("^\\.", states)) %>%
  mutate(states = gsub("^\\.", "", states))
```

## Data Background and Context

The dataset originates from the landmark study *"Systemic Discrimination Among Large U.S. Employers"* (Kline, Rose, and Walters, 2022).

-   **Research Questions:** The study explores whether discrimination is endemic to particular firms, investigates firm-level heterogeneity in callback rates, and considers the potential impact of industry, geographic location, and other structural factors.

The census data was retrieved from the United States Census Bureau at this link: https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html

# Aggregating Application Data to State Level
To begin our analysis, we aggregate individual-level application data into state-level summaries. In this section, we create new variables such as the full name (and its length) and calculate the proportion of applicants with black-sounding names. These measures help us capture nuances in the data that might be related to potential discriminatory practices.

```{r}
data$full_name <- paste(data$firstname, data$lastname)
data$name_length <- nchar(data$full_name)
state_data <- data %>%
  group_by(state, year) %>%
  summarise(
    num_sub = n(),
    callback_rate = mean(cb),
    age = mean(age_at_sub),
    name_length = mean(name_length),
    proportion_black_name = sum(black == 1) / num_sub
  ) %>%
  ungroup()
unique(state_data$state)
```

# Cleaning and Reshaping Census Data - continued

The next step involves cleaning the census data. This dataset provides annual state population counts, which we reshape into a longitudinal (panel) format to facilitate merging with the state-level application data. The cleaning process includes renaming columns, removing extraneous rows, and filtering to include only the relevant states.

```{r}
colnames(census_data)[1] <- "states"

census_data <- census_data %>% 
  select(-2) %>% 
  slice(-(1:3))

colnames(census_data)[2:6] <- c("2020", "2021", "2022", "2023", "2024")

census_data <- census_data %>% filter(grepl("^\\.", states))

census_data <- census_data %>% 
  filter(states != ".Puerto Rico") %>%
  mutate(states = gsub("^\\.", "", states)) 

# add census_data_2 to census_data 
census_data <- census_data %>%
  left_join(census_data_2, by = "states") 
```

# Clean Population Data to Longitudinal Format

```{r}
# create a state abbr column for merging
census_data$state_abbr <- toupper(substr(census_data$states, 1, 2))

# reshape the data into state-year panel
census_data_long <- census_data %>%
  pivot_longer(
    cols = -c(states, state_abbr),
    names_to = "year",
    values_to = "population",
    values_transform = list(population = as.numeric)
  ) %>%
  mutate(year = as.numeric(year))

# check unique state_abbr
unique(census_data$states)


```

```{r}
unique(state_data$year)
```

```{r}
unique(census_data_long$year)
```
We further process the census data to create a state-year panel, including the state abbreviations necessary for merging with the experimental data.
```{r}
state_mapping <- tibble::tribble(
  ~states,                     ~state_abbr_exp,
  "Alabama",                   "AL",
  "Alaska",                    "AK",
  "Arizona",                   "AZ",
  "Arkansas",                  "AR",
  "California",                "CA",
  "Colorado",                  "CO",
  "Connecticut",               "CT",
  "Delaware",                  "DE",
  "District of Columbia",      "DC",
  "Florida",                   "FL",
  "Georgia",                   "GA",
  "Hawaii",                    "HI",
  "Idaho",                     "ID",
  "Illinois",                  "IL",
  "Indiana",                   "IN",
  "Iowa",                      "IA",
  "Kansas",                    "KS",
  "Kentucky",                  "KY",
  "Louisiana",                 "LA",
  "Maine",                     "ME",
  "Maryland",                  "MD",
  "Massachusetts",             "MA",
  "Michigan",                  "MI",
  "Minnesota",                 "MN",
  "Mississippi",               "MS",
  "Missouri",                  "MO",
  "Montana",                   "MT",
  "Nebraska",                  "NE",
  "Nevada",                    "NV",
  "New Hampshire",             "NH",
  "New Jersey",                "NJ",
  "New Mexico",                "NM",
  "New York",                  "NY",
  "North Carolina",            "NC",
  "North Dakota",              "ND",
  "Ohio",                      "OH",
  "Oklahoma",                  "OK",
  "Oregon",                    "OR",
  "Pennsylvania",              "PA",
  "Rhode Island",              "RI",
  "South Carolina",            "SC",
  "South Dakota",              "SD",
  "Tennessee",                 "TN",
  "Texas",                     "TX",
  "Utah",                      "UT",
  "Vermont",                   "VT",
  "Virginia",                  "VA",
  "Washington",                "WA",
  "West Virginia",             "WV",
  "Wisconsin",                 "WI",
  "Wyoming",                   "WY"
)

census_data_long <- census_data_long %>%
  left_join(state_mapping, by = "states")

merged_data <- state_data %>%
  left_join(
    census_data_long %>% select(state_abbr_exp, year, population),
    by = c("state" = "state_abbr_exp", "year" = "year")
  )

# cross sectional data for all years all states
cross_sectional <- merged_data %>%
  group_by(state) %>%
  summarise(
    num_sub = sum(num_sub),
    callback_rate = mean(callback_rate),
    age = mean(age),
    name_length = mean(name_length),
    proportion_black_name = sum(proportion_black_name * num_sub) / num_sub,
    population = mean(population),
    proportion_of_applicants = num_sub / population
  ) %>%
  ungroup()


plot_usmap(data = cross_sectional, values = "population") +
  scale_fill_continuous(low = "white", high = "skyblue", 
                        name = "Population Size", 
                        label = scales::comma) +
  labs(title = "Mean Population Size of Each State 2019-2021") +
  theme(legend.position = "right")
 
```

```{r}
plot_usmap(data = cross_sectional, values = "num_sub") +
  scale_fill_continuous(low = "white", high = "skyblue", 
                        name = "Total Number of Applicantions", 
                        label = scales::comma) +
  labs(title = "Total Number of Applicantions Within State 2019-2021") +
  theme(legend.position = "right")
 
```
Here we can see that the relative number of applications across states actually reflect the size of the population in each state. 

Our analysis indicates that the number of applications in each state largely reflects its population size. In particular, states in the West, Northeast, and Southeast—regions known for having larger populations—consistently show higher application counts, aligning with the overall demographic distribution across the United States.


